---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# calibmsm

<!-- badges: start -->
[![R-CMD-check](https://github.com/alexpate30/calibmsm/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/alexpate30/calibmsm/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of calibmsm is to provide a set of tools for producing calibration plots for validating an existing (i.e. previously developed) multistate model.

## Installation

You can install the development version of calibmsm from [GitHub](https://github.com/) with:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("alexpate30/calibmsm")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(calibmsm)
## basic example code

## WRITE BASIC MINIMUM EXAMPLE
```

## Getting help
If you encounter a bug, please file an issue with a minimal reproducible example on [GitHub](https://github.com/alexpate30/calibmsm).

NB TO DELETE - A USEFUL NOTE FOR NOW:: You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. You could also use GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/v1/examples>.


