---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# calibmsm

<!-- badges: start -->
[![R-CMD-check](https://github.com/alexpate30/calibmsm/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/alexpate30/calibmsm/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/alexpate30/calibmsm/branch/master/graph/badge.svg)](https://app.codecov.io/gh/alexpate30/calibmsm/?branch=master)
<!-- badges: end -->

The goal of **calibmsm** is to provide a set of tools for estimating calibration plots when validating an existing (i.e. previously developed) multistate model.

## Installation

You can install the development version of calibmsm from [GitHub](https://github.com/) with:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("alexpate30/calibmsm")
```

## Example

This is a basic example which shows you how to assess the calibration of the transition probabilities at 5 years follow up for individuals out of state `j = 1` at time `s = 0` using the BLR-IPCW approach. Please see the *overview* vignette for examples of how to assess calibration using the pseudo-value and MLR-IPCW approaches, as well as details of the methodology. 

The predicted transition probabilities are stored in `tps0`, the individuals data are stored in `ebmtcal`, and the data in `msdata` format are stored in `msebmtcal`. Calibration curves are estimated using `calib_blr`. Inverse probability of censoring weights are calculated based on variables `year`, `age`, `prophylaxis` and donor gender `match`. The calibration curves are estimated using restricted cubic splines with 3 knots. A 95% confidence interval is calculated using bootstrapping with 200 bootstrap replicates.

```{r example}
## Load calibmsm
library(calibmsm)

## Extract relevant predicted risks from tps0
tp.pred <- dplyr::select(dplyr::filter(tps0, j == 1), any_of(paste("pstate", 1:6, sep = "")))

## Calculate observed event probabilities
dat.calib.blr <-
  calib_blr(data.mstate = msebmtcal,
                 data.raw = ebmtcal,
                 j=1,
                 s=0,
                 t = 1826,
                 tp.pred = tp.pred,
                 curve.type = "rcs",
                 rcs.nk = 3,
                 w.covs = c("year", "agecl", "proph", "match"),
                 CI = 95,
                 CI.R.boot = 200)

## Produce summary
summary(dat.calib.blr)

## Plot calibration plots
plot(dat.calib.blr, combine = TRUE, nrow = 2, ncol = 3)
```

## Getting help
If you encounter a bug, please file an issue with a minimal reproducible example on [GitHub](https://github.com/alexpate30/calibmsm).


