---
title: "Evaluation-of-estimation-of-IPCWs"
output: rmarkdown::html_vignette
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{Evaluation-of-estimation-of-IPCWs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

When assessing calibration using the BLR-IPCW and pseudo-value approaches (see the `vignette("[Overview]")` vignette), we found predominately good agreement between the calibration curves except when assessing the calibration of the transition probabilities from state $j = 1$ into state $k = 3$. Given we do not know the true calibration of these transition probabilities we cannot say for sure which calibration curve is correct. However, we can take steps to test the assumptions made by each of the methods, which may help guide which method of assessing calibration should be used in this particular clinical example.

# Methods and Results

BLR-IPCW requires that the outcome, $I_{k}(t)$, is independent from the censoring mechanism in the re-weighted population. Psuedo-values require non-informative censoring within the subgroups within which they are calculated. A key variable which predicts the censoring mechanism is year of transplant,  `year`. Individuals who had their transplant more recently have a shorter administrative censoring time, given they have a shorter maximum follow up. 

When calculating pseudo-values, we can completely remove the impact of this variable by calculating pseudo-values within subgroups defined by this variable. In the `vignette("[Overview]")` vignette, we also further group patients by their predicted risk in a further three subgroups. We therefore believe the assumption for the pseudo-value approach is likely to hold.

When implementing the BLR-IPCW approach, the most important step is estimation of the weights. In `calibmsm` this is done internally using a cox proportional hazards model. We fit this model to the `ebmtcal` data manually to assess how well it has been specified. 

We start by loading the `ebmtcal` data.

```{r}
set.seed(101)

library(calibmsm)

data("ebmtcal")
head(ebmtcal)
```

This is followed by fitting a cox model to predict probability of being censored. This is the model used to estimate the inverse probability of censoring weights (IPCWs) when implementing the BLR-IPCW method using `calib_blr` and setting `w.covs = c("year", "agecl", "proph", "match")`.

```{r}
model.ipcw <- survival::coxph(survival::Surv(dtcens, dtcens.s) ~ year + agecl + proph + match, data = ebmtcal)
model.ipcw
```

The coefficient for `year`, in particular 1995 - 1998, is very big. However, we know that year of transplant is likely be to highly predictive of being censored, so further exploration is needed. A key assumption made by this model is proportional hazards. We plot the the cumulative hazard estimate using Kaplan-Meier survival curves stratified by year of transplant and assess proportionality visually.

```{r}
kmfit <- survival::survfit(survival::Surv(dtcens, dtcens.s) ~ year, data = ebmtcal)
```

```{r, fig.width = 7.5, fig.height = 5, fig.fullwidth = TRUE, fig.cap = "Figure 1: Cumulative hazard calculated within subgroups defined by year of transplant", fig.topcaption = TRUE}
plot(kmfit, fun = "cumhaz", col = c(1,2,3), xlab = "Days", ylab = "Cumulative hazard")
legend(1,6.5,c("1985 - 1989", "1990 - 1994", "1995 - 1998"), col = c(1,2,3), lty = c(1, 1, 1))
```

# Discussion

Clearly the proportional hazards assumption does not hold. This means the model for estimating the weights is misspecified and the estimated IPCWs are likely to be wrong. While the variable year of transplant could be removed from the model for estimating the weights to help ensure the proportional hazards assumption is met, we know this variable is highly predictive of being censored, and therefore omitting this key predictor would also result in a misspecified model and incorrect weights. We would therefore conclude that it is unlikely that the assumptions hold for the BLR-IPCW method, which could be resulting in biased calibration curves in this particular scenario. If developing this model in practice, we would recommend assessing calibration using the pseudo-value approach. This is not a fundamental problem with the BLR-IPCW approach, but an issue with estimating the IPCWs, which if estimated correctly should result in unbiased calibration curves. As suggested in the `vignette("[Overview]")` vignette, more simulation studies are required to rigorously test the robustness of each method when their assumptions do not hold, and in what scenario each method is most likely to give biased calibration curves.

This vignette has also highlighted the limitations of using a cox proportional hazards model to estimate the IPCWs. The `calibmsm` package has been developed to provide a set of tools to assess the calibration of multistate models, but is not a focused package for estimating IPCWs. We urge users to develop better models for estimating the IPCWs and specify them using the `weights` argument. A process for estimating confidence intervals using bootstrapping when manually estimating the IPCWs is exemplified in the vignette `vignette("[BLR-IPCW-manual-boostrap]")`.

One final point, it is peculiar that this issue only causes discordance between the calibration plots out of state $j = 0$ into state $k = 3$, whereas for all other states $k$ there is closer agreement between the BLR-IPCW and pseudo-value calibration curves (see Figures 2 and 3 in the `vignette("[Overview]")` vignette). We hypothesise that this may be due to the fact no individuals move into state $3$ after $100$ days. We believe this is likely due to the definition of an adverse event needing to happen within $100$ days post transplant, but cannot be sure as the authors of this package had no involvement in data collection. It is possible that the application of incorrectly estimated weights, in combination with predicted risks of the outcome state which behaves differently to all others, is the reason behind this behaviour. However, further exploration of this is beyond the scope of this vignette.
