---
title: "Calibration-curves-estimated-with-loess-smoothers"
output: rmarkdown::html_vignette
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{Calibration-curves-estimated-with-loess-smoothers}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Data preperation

This vignette showcases how to estimate BLR-IPCW and pseudo-value calibration curves when using loess smoothers for the model to estimate the observed event probabilities. This is in contrast to using restricted cubic splines, which was done so in the [Overview vignette](https://alexpate30.github.io/calibmsm/articles/overview_vignette.pdf). Both approaches are viable and we refer to the literature for a discussion of each method [@Austin2020; @Harrell2015]. We mimic exactly the analysis from the [Overview vignette](https://alexpate30.github.io/calibmsm/articles/overview_vignette.pdf), specifying loess smoother rather than restricted cubic splines, and assess whether the conclusions would be any different. This is primarily to showcase the utility of the package.

We use data from the European Society for Blood and Marrow Transplantation [@EBMT2023], which contains multistate survival data after a transplant for patients with blood cancer. The start of follow up is the day of the transplant and the initial state is alive and in remission. There are three intermediate events ($2$: recovery, $3$: adverse event, or $4$: recovery + adverse event), and two absorbing states ($5$: relapse and $6$: death). This data was originally made available from the `mstate` package [@DeWreede2011].

We start by reminding ourselves of the EBMT [@DeWreede2011; @EBMT2023] validation datasets `ebmtcal` and `msebmtcal`, and the predicted transition probabilities out of state `j = 1` made at time `s = 0` from the multistate model (`tps0`). We also set the time at which we want to evaluate the transition probabilities `t`. Please refer to the [Overview vignette](https://alexpate30.github.io/calibmsm/articles/overview_vignette.pdf) for a more detailed description of this data.

```{r}
set.seed(101)

library(calibmsm)

data("ebmtcal")
head(ebmtcal)

data("msebmtcal")
head(msebmtcal)

data("tps0")
head(tps0)

t.eval <- 1826
```

# Estimation of calibration curves out of state j = 1 at time s = 0 using loess smoothers

We now produce calibration curves for the predicted transition probabilities out of state $j = 1$ at time $s = 0$. Given all individuals start in state $1$, there is no need to consider the transition probabilities out of states $j \neq 1$ at $s = 0$. Calibration is assessed at follow up time ($t = 1826$ days). We start by extracting the predicted transition probabilities from state $j = 1$ at time $s = 0$ from the object \code{tps0}. These are the transition probabilities we aim to assess the calibration of.

```{r}
tp.pred.s0 <- tps0 |>
  dplyr::filter(j == 1) |>
  dplyr::select(any_of(paste("pstate", 1:6, sep = "")))
```

We now produce calibration curves using loess smoothers for the regression model where the observed event probabilities are estimated. This is specified by \code{curve.type = "loess"}. Behaviour of the curve can be controlled using \code{loess.span} and \code{loess.degree}, which we leave as default in this example. Please refere to documentation for the \code{loess} function from the \pkg{loess} package for details of these parameters. A smaller span and/or degree will create a more smoother line. This is done for both the BLR-IPCW and pseudo-value calibration curves.

```{r}
dat.calib.blr <-
  calib_msm(data.ms = msebmtcal,
           data.raw = ebmtcal,
           j = 1,
           s = 0,
           t = t.eval,
           tp.pred = tp.pred.s0,
           calib.type = 'blr',
           curve.type = "loess",
           w.covs = c("year", "agecl", "proph", "match"),
           CI = 95,
           CI.type = "bootstrap",
           CI.R.boot = 200)

dat.calib.pv <-
  calib_msm(data.ms = msebmtcal,
           data.raw = ebmtcal,
           j = 1,
           s = 0,
           t = t.eval,
           tp.pred = tp.pred.s0,
           calib.type = 'pv',
           curve.type = "loess",
           pv.group.vars = c("year"),
           pv.n.pctls = 3,
           CI = 95,
           CI.type = 'parametric')
```

We now produce calibration plots using the \code{plot} function.

```{r, fig.width = 7.5, fig.height = 5, fig.fullwidth = TRUE, fig.cap = "Figure 1: BLR-IPCW calibration plots out of state j = 1 at time s = 0  estimated using loess smoothers"}
plot(dat.calib.blr)
```

```{r, fig.width = 7.5, fig.height = 5, fig.fullwidth = TRUE, fig.cap = "Figure 2: pseudo-value calibration plots out of state j = 1 at time s = 0 estimated using loess smoothers"}
plot(dat.calib.pv)
```

The equivalent Figures from the [Overview vignette](https://alexpate30.github.io/calibmsm/articles/overview_vignette.pdf), produced using restricted cubic splines, are Figures 2 and 3.

Write a short comparison... XXXX

# Estimation of calibration curves out of state j = 1 at time s = 100 using loess smoothers

We now produce calibration curves for the predicted transition probabilities out of state $j = 1$ at time $s = 100$. Calibration is assessed at follow up time ($t = 1826$ days). We start by extracting the predicted transition probabilities from state $j = 1$ at time $s = 100$ from the object \code{tps0}. These are the transition probabilities we aim to assess the calibration of.

```{r}
tp.pred.s100 <- tps100 |>
  dplyr::filter(j == 1) |>
  dplyr::select(any_of(paste("pstate", 1:6, sep = "")))
```

We now produce calibration curves using loess smoothers for the regression model where the observed event probabilities are estimated. This is specified by \code{curve.type = "loess"}. Behaviour of the curve can be controlled using \code{loess.span} and \code{loess.degree}, which we leave as default in this example. Please refere to documentation for the \code{loess} function from the \pkg{loess} package for details of these parameters. A smaller span and/or degree will create a more smoother line. This is done for both the BLR-IPCW and pseudo-value calibration curves.

```{r}
dat.calib.blr <-
  calib_msm(data.ms = msebmtcal,
           data.raw = ebmtcal,
           j = 1,
           s = 100,
           t = t.eval,
           tp.pred = tp.pred.s100,
           calib.type = 'blr',
           curve.type = "loess",
           w.covs = c("year", "agecl", "proph", "match"),
           CI = 95,
           CI.type = "bootstrap",
           CI.R.boot = 200)

dat.calib.pv <-
  calib_msm(data.ms = msebmtcal,
           data.raw = ebmtcal,
           j = 1,
           s = 100,
           t = t.eval,
           tp.pred = tp.pred.s100,
           calib.type = 'pv',
           curve.type = "loess",
           pv.group.vars = c("year"),
           CI = 95,
           CI.type = 'parametric')
```

We now produce calibration plots using the \code{plot} function.

```{r, fig.width = 7.5, fig.height = 5, fig.fullwidth = TRUE, fig.cap = "Figure 3: BLR-IPCW calibration plots out of state j = 1 at time s = 100 estimated using loess smoothers"}
plot(dat.calib.blr)
```

```{r, fig.width = 7.5, fig.height = 5, fig.fullwidth = TRUE, fig.cap = "Figure 4: pseudo-value calibration plots out of state j = 1 at time s = 100 estimated using loess smoothers"}
plot(dat.calib.pv)
```

The equivalent Figures from the [Overview vignette](https://alexpate30.github.io/calibmsm/articles/overview_vignette.pdf), produced using restricted cubic splines, are Figures 5 and 6.

Write a short comparison...

# Estimation of calibration curves out of state j = 3 at time s = 100 using loess smoothers

We now produce calibration curves for the predicted transition probabilities out of state $j = 3$ at time $s = 100$. Calibration is assessed at follow up time ($t = 1826$ days). We start by extracting the predicted transition probabilities from state $j = 3$ at time $s = 100$ from the object \code{tps0}. These are the transition probabilities we aim to assess the calibration of.

```{r}
tp.pred.s100 <- tps100 |>
  dplyr::filter(j == 3) |>
  dplyr::select(any_of(paste("pstate", 1:6, sep = "")))
```

We now produce calibration curves using loess smoothers for the regression model where the observed event probabilities are estimated. This is specified by \code{curve.type = "loess"}. Behaviour of the curve can be controlled using \code{loess.span} and \code{loess.degree}, which we leave as default in this example. Please refere to documentation for the \code{loess} function from the \pkg{loess} package for details of these parameters. A smaller span and/or degree will create a more smoother line. This is done for both the BLR-IPCW and pseudo-value calibration curves.

```{r}
dat.calib.blr <-
  calib_msm(data.ms = msebmtcal,
           data.raw = ebmtcal,
           j = 3,
           s = 100,
           t = t.eval,
           tp.pred = tp.pred.s100,
           calib.type = 'blr',
           curve.type = "loess",
           w.covs = c("year", "agecl", "proph", "match"),
           CI = 95,
           CI.type = "bootstrap",
           CI.R.boot = 200)

dat.calib.pv <-
  calib_msm(data.ms = msebmtcal,
           data.raw = ebmtcal,
           j = 3,
           s = 100,
           t = t.eval,
           tp.pred = tp.pred.s100,
           calib.type = 'pv',
           curve.type = "loess",
           pv.group.vars = c("year"),
           CI = 95,
           CI.type = 'parametric')
```

We now produce calibration plots using the \code{plot} function.

```{r, fig.width = 7.5, fig.height = 5, fig.fullwidth = TRUE, fig.cap = "Figure 5: BLR-IPCW calibration plots out of state j = 3 at time s = 100 estimated using loess smoothers"}
plot(dat.calib.blr)
```

```{r, fig.width = 7.5, fig.height = 5, fig.fullwidth = TRUE, fig.cap = "Figure 6: pseudo-value calibration plots out of state j = 3 at time s = 100 estimated using loess smoothers"}
plot(dat.calib.pv)
```

The equivalent Figures from the [Overview vignette](https://alexpate30.github.io/calibmsm/articles/overview_vignette.pdf), produced using restricted cubic splines, are Figures 7 and 8.

Write a short comparison... XXXX

# A note on the fact that confidence intervals may be lower than 0 or bigger than 1

Given the standard error is calculated on the probability scale here, as opposed to the logit scale, the confidence intervals may exceed 0 and 1. I have checked and the confidence intervals are calculated in the same way in the CalibrationCurves package: https://cran.r-project.org/web/packages/CalibrationCurves/CalibrationCurves.pdf . In particular see code for val.prob.ci.2: https://github.com/BavoDC/CalibrationCurves/blob/master/R/val.prob.ci.2.R . There parametric confidence intervals are calculated in the same way we calculate parametric confidence intervals for the pseudo value approach.  I am therefore confident this is correct, but is something i'd like to come back to and consider the implications of. It's also intriguing that even for the bootstrapped BLR-IPCW confidence intervals, we are seeing confidence intervals well outside 0 and 1. I thought this would be an issue for the pseudo-value parametric confidence intervals as we are just adding on 1.96*se on the probability scale. However, I didnt actually think that would be possible for the bootstrapped confidence intervals, as in any given bootstrapped dataset, predicted-observed values would still lie between 0 and 1, so this is something to explore in more detail.

Having done a bit more research (see test_loess_bootstrap.R in the test folder), the values below 0 are driven by a small number of individuals with the smallest risks. Lots of their predicted-observed values are NAs, and they have some negative values too. It is evident these small number of individuals are driving all the NA values we are finding in each bootstrap iteration. Going to try increasing the span to see if that helps. Now i know the mechanism, although not something I'm particularly worried about, as seems to be driven by small sample size when assessing calibration at time s = 100. 

Compare the following graphs to figures 1 and 2 above. First do \code{loess.span = 0.9}.

```{r}
dat.calib.blr <-
  calib_msm(data.ms = msebmtcal,
           data.raw = ebmtcal,
           j = 1,
           s = 0,
           t = t.eval,
           tp.pred = tp.pred.s0,
           calib.type = 'blr',
           curve.type = "loess",
           loess.span = 0.9,
           w.covs = c("year", "agecl", "proph", "match"),
           CI = 95,
           CI.type = "bootstrap",
           CI.R.boot = 200)

dat.calib.pv <-
  calib_msm(data.ms = msebmtcal,
           data.raw = ebmtcal,
           j = 1,
           s = 0,
           t = t.eval,
           tp.pred = tp.pred.s0,
           calib.type = 'pv',
           curve.type = "loess",
           loess.span = 0.9,
           pv.group.vars = c("year"),
           pv.n.pctls = 3,
           CI = 95,
           CI.type = 'parametric')
```

We now produce calibration plots using the \code{plot} function.

```{r, fig.width = 7.5, fig.height = 5, fig.fullwidth = TRUE, fig.cap = "Figure 7: BLR-IPCW calibration plots out of state j = 1 at time s = 0  estimated using loess smoothers, span = 0.9"}
plot(dat.calib.blr)
```

```{r, fig.width = 7.5, fig.height = 5, fig.fullwidth = TRUE, fig.cap = "Figure 8: pseudo-value calibration plots out of state j = 1 at time s = 0 estimated using loess smoothers, span = 0.9"}
plot(dat.calib.pv)
```

Then do \code{loess.span = 1}.

```{r}
dat.calib.blr <-
  calib_msm(data.ms = msebmtcal,
           data.raw = ebmtcal,
           j = 1,
           s = 0,
           t = t.eval,
           tp.pred = tp.pred.s0,
           calib.type = 'blr',
           curve.type = "loess",
           loess.span = 1,
           w.covs = c("year", "agecl", "proph", "match"),
           CI = 95,
           CI.type = "bootstrap",
           CI.R.boot = 200)

dat.calib.pv <-
  calib_msm(data.ms = msebmtcal,
           data.raw = ebmtcal,
           j = 1,
           s = 0,
           t = t.eval,
           tp.pred = tp.pred.s0,
           calib.type = 'pv',
           curve.type = "loess",
           loess.span = 1,
           pv.group.vars = c("year"),
           pv.n.pctls = 3,
           CI = 95,
           CI.type = 'parametric')
```

We now produce calibration plots using the \code{plot} function.

```{r, fig.width = 7.5, fig.height = 5, fig.fullwidth = TRUE, fig.cap = "Figure 9: BLR-IPCW calibration plots out of state j = 1 at time s = 0  estimated using loess smoothers, span = 1"}
plot(dat.calib.blr)
```

```{r, fig.width = 7.5, fig.height = 5, fig.fullwidth = TRUE, fig.cap = "Figure 10: pseudo-value calibration plots out of state j = 1 at time s = 0 estimated using loess smoothers, span = 1"}
plot(dat.calib.pv)
```

