---
title: "Evaluation-of-estimation-of-IPCWs"
output: rmarkdown::html_vignette
bibliography: refs.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette explores whether the assumptions of the BLR-IPCW and psuedo-value methods for assessing calibration are met in the illustrative example in the [Overview vignette](https://alexpate30.github.io/calibmsm/articles/overview_vignette.pdf). We use data from the European Society for Blood and Marrow Transplantation [@EBMT2023], which contains multistate survival data after a transplant for patients with blood cancer. The start of follow up is the day of the transplant and the initial state is alive and in remission. There are three intermediate events ($2$: recovery, $3$: adverse event, or $4$: recovery + adverse event), and two absorbing states ($5$: relapse and $6$: death). This data was originally made available from the `mstate` package [@DeWreede2011].

In the illustrative example in the [Overview vignette](https://alexpate30.github.io/calibmsm/articles/overview_vignette.pdf), we found predominately good agreement between the BLR-IPCW and pseudo-value calibration curves, except when assessing the calibration of the transition probabilities from state $j = 1$ into state $k = 3$. Given we do not know the true calibration of these transition probabilities we cannot say for sure which calibration curve is correct. However, we can take steps to test the assumptions made by each of the methods, which may help guide which method of assessing calibration should be used in this particular clinical example. We suggest package users take similar steps to explore which method may be most appropriate in their particular clinical context.

# Methods and Results

We start by reminding ourselves of the EBMT [@DeWreede2011; @EBMT2023] validation dataset `ebmtcal`. Please refer to the [Overview vignette](https://alexpate30.github.io/calibmsm/articles/overview_vignette.pdf) for a more detailed description of this data.

BLR-IPCW requires that the outcome, $I_{k}(t)$, is independent from the censoring mechanism in the re-weighted population. Psuedo-values require non-informative censoring within the subgroups within which they are calculated. A key variable which predicts the censoring mechanism is year of transplant,  `year`. Individuals who had their transplant more recently have a shorter administrative censoring time, given they have a shorter maximum follow up. 

When calculating pseudo-values, we can completely remove the impact of this variable by calculating pseudo-values within subgroups defined by this variable. In the [Overview vignette](https://alexpate30.github.io/calibmsm/articles/overview_vignette.pdf) we also further grouped patients by their predicted risk in a further three subgroups. We therefore believe the assumption for the pseudo-value approach is likely to hold.

When implementing the BLR-IPCW approach, the most important step is estimation of the inverse probability of censoring weights (IPCWs). In `calibmsm` this is done internally using a cox proportional hazards model. We fit this model to the `ebmtcal` data again here to assess how well it has been specified. 

We start by loading the `ebmtcal` data.

```{r}
set.seed(101)

library(calibmsm)

data("ebmtcal")
head(ebmtcal)
```

This is followed by fitting a cox model to predict probability of being censored. This below code chunk is what happens internally in `calib_blr` if one sets `w.covs = c("year", "agecl", "proph", "match")`.

```{r}
model.ipcw <- survival::coxph(survival::Surv(dtcens, dtcens.s) ~ year + agecl + proph + match, data = ebmtcal)
model.ipcw
```

The coefficient for `year`, in particular 1995 - 1998, is very big. However, we know that year of transplant is likely be to highly predictive of being censored, so further exploration is needed. A key assumption made by this model is proportional hazards. We plot the the cumulative hazard estimate using Kaplan-Meier survival curves stratified by year of transplant and assess proportionality visually.

```{r}
kmfit <- survival::survfit(survival::Surv(dtcens, dtcens.s) ~ year, data = ebmtcal)
```

```{r, fig.width = 7.5, fig.height = 5, fig.fullwidth = TRUE, fig.cap = "Figure 1: Cumulative hazard calculated within subgroups defined by year of transplant", fig.topcaption = TRUE}
plot(kmfit, fun = "cumhaz", col = c(1,2,3), xlab = "Days", ylab = "Cumulative hazard")
legend(1,6.5,c("1985 - 1989", "1990 - 1994", "1995 - 1998"), col = c(1,2,3), lty = c(1, 1, 1))
```

# Discussion

For the proportional hazards assumption to hold, the Kaplan-Meier survival curves within in each group should be proprtional to eachother over every follow up time. Clearly the proportional hazards assumption does not hold. This means the model for estimating the weights is misspecified and the estimated IPCWs may be wrong. While the variable year of transplant could be removed from the model for estimating the weights to help ensure the proportional hazards assumption is met, we know this variable is highly predictive of being censored, and therefore omitting this key predictor would also result in a misspecified model and incorrect weights. 

We would therefore conclude that it is unlikely that the assumptions hold for the BLR-IPCW method, which could be resulting in biased calibration curves in this particular clinical example. If developing this model in practice, we would recommend assessing calibration using the pseudo-value approach. This is not a fundamental problem with the BLR-IPCW approach, but an issue with estimating the IPCWs under the censoring mechanism in this example. If the IPCWs can be estimated correctly, this should result in unbiased calibration curves. As suggested in the [Overview vignette](https://alexpate30.github.io/calibmsm/articles/overview_vignette.pdf), more simulation studies are required to rigorously test the robustness of each method when their assumptions do not hold, and in what scenario each method is most likely to give biased calibration curves.

This vignette has also highlighted the limitations of using a cox proportional hazards model to estimate the IPCWs. The `calibmsm` package has been developed to provide a set of tools to assess the calibration of multistate models, but is not a focused package for estimating IPCWs. We urge users to develop better models for estimating the IPCWs and specify them using the `weights` argument. A process for estimating confidence intervals using bootstrapping when manually estimating the IPCWs is exemplified in the [BLR-IPCW-manual-boostrap vignette](BLR-IPCW-manual-boostrap.html).

One final point, it is peculiar that this issue only causes discordance between the calibration plots out of state $j = 0$ into state $k = 3$, whereas for all other states $k$ there is closer agreement between the BLR-IPCW and pseudo-value calibration curves (see Figures 2 and 3 in the [Overview vignette](https://alexpate30.github.io/calibmsm/articles/overview_vignette.pdf)). We hypothesise that this may be due to the fact no individuals move into state $3$ after $100$ days. We believe this is due to the definition of an adverse event needing to happen within $100$ days post transplant, but cannot be sure as the authors of this package had no involvement in data collection. It is possible that the application of incorrectly estimated weights, in combination with predicted risks of the outcome state which behaves differently to all others, is the reason behind this behaviour. However, further exploration of this is beyond the scope of this vignette.
